Understanding the technological transition based on cultural transmission biases during the Late Pleistocene in Korea

### 2021-02-01 analysis/paper/paper.Rmd
moved writing related texts to my google notes

# Introduction

1. What are our research questions?
- Understanding technological transition during the Late Pleistocene in Korea base on Cultural transmission. 
- What are the social contexts led to the technological transition

2. What is our archaeological background?
-SP was started in Korean Peninsula -> spread to Japan
-Why SP important, Why we care
-Current explantion of SP dispersal: migration vs in situ, in between
 The evidence of the two main models: proportion of SP in assemblage, Y chromosome (two groups of people), composition of assemblage (blade to flake ratio)
 -> no explanation about any mechanism/process/driver for the new technology
 -> lack of details of the lithic technology, human behavior

3. What is our conceptual framework?
- Evolutionary concepts that links stone artifacts to human behavior
- Cultural Evolution : the idea that cultural change constitutes an evolutionary process
- Same general concept of biological evolution: genetic change in population that is inherited over generations. Apply this concept to artifacts or cultural phenomena

- In this paper, I will use Cultural Transmission 
- Cultural Transmssion: way that people learn and pass on information
- Transmission biases
- similar study (CT in stone artifacts) : Mesoudi&O'Brien, Bettinger&Eerkens, Garvey

-------------
Application the theoretical and methodological tenets of Darwinian evolutionary concpet to archaeological studies has helped understand technological transition related to human behavior (Dunnell 1980; Lipo et al. 1997; Bettinger and Eerkens 1997, 1999, 2008, Mesoudi abd O'Brien 2008).

biological concept to archaeological usage: (Richerson and Boyd 1992)

Cultrual evolution that archaeologists adapted has three sub-categories: human behavioral ecology, phylogenetics and cultural transmission theory (Garvey 2018). Among them, we apply models and methods of cultural tranmssion to understand the technological change and socio cultural contexts.

Cultural transmission is the idea that a transition can occur during the process of social learning and the transmission biases are important loci of changes in material cultural (Lycett 2015; Creanza et al. 2017). In other words, this concept holds that instructions for making artifacts are transmitted between members of a group during social learning by observation of, or interaction with another member (Heyes 1994; Eerkens and Lipo 2007; Brantingham et al. 2010; Lycett 2015).

The transmission is subject to biases that result from different social contexts. Boyd and Richerson (1985) and Henrich and McElreath (2003) have defined many of these biases, including ‘guided variation’ (where individuals learn new behaviors and then modify them through trial and error), ‘content-based bias’ (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), ‘frequency-based bias’ (where an individual is biased to choose particular instructions based on their perceived frequency in the population, such as extremely popular or rare instructions), and ‘model-based bias’ (where a variant is transmitted because of its association with other attributes, such as the prestige or skill of other individuals).

Bettinger and Eerkens (1997; 1999; 2008) have shown how these biases can be identified in stone artifact assemblages. They studied metric variables of stone points during the introduction of bow and arrow technology in the Great Basin around AD 300-600. They equated guided transmission with high metric variation and low correlation between metric attributes, and equated model-based bias with less variation and more correlated variables. They found that in eastern California artifacts have poor correlations of basal width and mass, suggesting new technology was introduced and spread by guided transmission.On the other hand, in central Nevada metric variables are highly correlated, indicating that the bow and arrow was introduced and spread by model-based bias. Since Bettiniger and Eerkens studied archaeological data so they are unable to directly observe the transmission process. To test the underlying assumptions that Bettinger and Eerkens used, Mesoudi and O’Brien (2008) have conducted experiments that simulated prestige bias and guided variation, and observed results that supported Bettinger and Eerkens’ hypothesis.

Garvey

-------------

4. What are our methods?
- Framework: Bettinger&Eerkens
- Two models: Guided variation and Model-based bias =>  Multi-point Origin (in situ or in between one) and Single-point Origin (migration model)
- Test by Correlation and Variation (Pearson's r, PCA, Coefficient of Variation) 
- data collection, sites, artifact #
- attributes: landmarks (tps software)

5. What are our results? 
- overall change through time
- correlation: postively correlated(?). We are using PCA too
- PCA: 2,3 phases are highly overlapped, high redundancy. The individual phase are scattered, low redundancy. Variables of Dimension 1 is positively correlated
- CV:values are around 20(ML)-34(SL). Compared to 2nd phase, 3rd phase has slightly lower variation.

6. What is the answer to our questions??
- low correlation + high variation? High correlation + low variation?
- What are the criteria for correlation and variation
- How was SP spread in Korea, did they have single orgin or multiple origins?
- What are the impact of combinaton with blade technology

7. What is limitation of this study?
- Maybe applying CV for SP is not ideal. It can be depending on individual knapper's skill and raw materials
- CV of one site (SYG6) - no big difference with the value of all data
- CV: lithis vs pottery

- couldn't include thickness and weight

** Expectation
- The main measurements are correlation(+PCA) and CV.
- The set of the two measurements would be like: 
1) lower CV and higher correlation between sites, in this case, we will think that SPs were the result of model based bias and possibly had single origin. 
2) higher CV and lower correlation between sites, for this case, we will thank that SPs were the result of guided variation and possibly had multiple origins. 
- We have three phases, but there are only two artifacts for the first phase. There are more artifacts and sites for the first phase but not eligible for this research because we gathered complete or almost complete shape of SPs. We excluded broken pieces. Therefore, we focus on the change from the 2nd phase to 3rd phase for understanding the process of SP's dispersal in the Korean Peninsula.
- Change from the 2nd to 3rd
1) less correlated: the two groups were not very related/locally developed 
2) highly correlated: the two groups were related, had single source and then spread to other places
- Site by site
- Regional relationship? Province/cities

# Background

# Methods

Reading tps data and phase information of each artifact

@@

@@

library(tidyr)
cv_plot_site %>% 
pivot_longer(cols = -sitename, names_to = "group") %>% 
ggplot(aes(x = group, y = value, fill = as.factor(sitename))) + 
geom_bar(stat="identity", position=position_dodge())

@@

library(tidyr)
cv_plot_stage %>% 
pivot_longer(cols = -SPstage.Stage, names_to = "group") %>%
ggplot(aes(x = group, y = value, fill = as.factor(SPstage.Stage))) + 
geom_bar(stat="identity", position=position_dodge()) + labs(x="Attributes", fill= "Phase")
```

@@

cv_by_site_df_label_more_than_10 %>%
  pivot_longer(cols = -c(site_phase, label), 
               names_to = "lithic_attribute",
               values_to = "CV") %>%
  ggplot(aes(x = lithic_attribute, 
             y = CV)) +
  geom_col() +
  facet_wrap(~  label) +
  ggtitle("CVs of point attributes by site (with sites >10 artefacts")

cv_by_site_df_label_more_than_10_ml_mw_tw <- 
cv_by_site_df_label_more_than_10 %>%
  pivot_longer(cols = -c(site_phase, label), 
               names_to = "lithic_attribute",
               values_to = "CV") %>%
  filter(lithic_attribute %in% c("ML", "MW", "TW"))  %>% 
  mutate(phase = parse_number(str_extract(site_phase, "_."))) %>% 
  group_by(site_phase) %>% 
  mutate(mean_cv = mean(CV))

@@

>>>>

@@

Attributes that we are using here:

It must be noted that all of this previously undertaken work has focused exclusively on the size (i.e., scale) of artifactual attributes and the patterns of variation that may be produced via copying error rates due to limits in size perception. However, variation in the overall shape of artifacts is also evidently important, given that aspects of shape may have specific functional or aesthetic properties (Schillinger et al. 2014).

what other people use:

@@

library(tidyr)
cv_plot_site %>% 
pivot_longer(cols = -sitename, names_to = "group") %>% 
ggplot(aes(x = group, y = value, fill = sitename)) + 
geom_bar(stat="identity", position=position_dodge())

@@

library(tidyr)
cv_plot_stage %>% 
pivot_longer(cols = -SPstage.Stage, names_to = "group") %>%
ggplot(aes(x = group, y = value, fill = as.factor(SPstage.Stage))) + 
geom_bar(stat="identity", position=position_dodge()) + labs(x="Attributes", fill= "Phase")

@@

cv_by_site_df_label_more_than_10 %>%
  pivot_longer(cols = -c(site_phase, label), 
               names_to = "lithic_attribute",
               values_to = "CV") %>%
  ggplot(aes(x = lithic_attribute, 
             y = CV)) +
  geom_col() +
  facet_wrap(~  label) +
  ggtitle("CVs of point attributes by site (with sites >10 artefacts)")

cv_by_site_df_label_more_than_10_ml_mw_tw <- 
cv_by_site_df_label_more_than_10 %>%
  pivot_longer(cols = -c(site_phase, label), 
               names_to = "lithic_attribute",
               values_to = "CV") %>%
  filter(lithic_attribute %in% c("BL", "ML", "MW", "TW"))  %>% 
  mutate(phase = parse_number(str_extract(site_phase, "_."))) %>% 
  group_by(site_phase) %>% 
  mutate(mean_cv = mean(CV))

@@

Eerkens and Lipo (2005) demonstrated that the thickness of Rose Springs projectile points from Owens Valley (USA) varied according to the predictions of the random accumulated copying error model. However, the basal width of these projectile points exhibited less variation than expected, suggesting that some non-mutation process (i.e., some form of stabilizing selection such as conformity) was operating on this particular variable.
